---
title: "Group Project Code"
author: "Madeline, Maisie, Audrey"
format: pdf
date: "2026-05-11"
---

# 1. Set up

## Packages
```{r}
#| label: reading-in-packages
#| message: false

# general use
library(tidyverse)
library(janitor)

# file organization
library(here)

# Boxplot Visualization
#install.packages("hrbrthemes")
library(hrbrthemes)
#install.packages("viridis")
library(viridis)

# Ridgeline Plot
#install.packages("ggridges")
library(ggridges)
#install.packages("ggplot2")
library(ggplot2)

# Color Palette
library(NatParksPalettes)
```

## Data
```{r}
#| label: data
#| message: false

# weather data from NOAA weather station
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))

```

## Code

```{r}
#| label: cleaning-weather

weather_clean <- weather |> 
  # made column names all lowercase
  clean_names() |> 
  # puts date in proper format and makes sure its understood as date (not as chr)
  mutate(date = mdy(date)) |> 
  # keeping only these 4 columns of interest
  select(date, prcp, tmax, tmin) |> 
  # creating new columns for month and year
  # extracting month from the date (needs to be in the standard date format for this function to work)
  mutate(month = month(date),
         year = year(date)) |> 
  mutate(wy = case_when(
    # when month os october or later, assign a value of year +1
    # in the water ywar for next year
    month >= 10 ~ year + 1,
    # keeping all months < 10 the same (sept or earlier), assigning value of existing year
    .default = year
    ))

```

```{r}
#| label: cleaning-metadata

metadata_clean <- metadata |> 
  # col names easier to work with (all lowercase w/ _)
  clean_names() |> 
  # "!" is the "not-operator", columns we DONT want to keep, excluding
  select(!c(object_id, creator, editor)) |> 
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  rename(monitoring_datetime = monitoring_date) |> 
  mutate(month = month(monitoring_datetime),
         year = year(monitoring_datetime)) |> 
  # assigning water year
  mutate(wy = case_when(
    month >= 10 ~ year +1,
    .default = year )) |> 
  # creatig new column for site full name
  mutate(site_full = case_when(
    site_name == "east_channel" ~ "East Channel",
    site_name == "phelps_bridge" ~ "Phelps Bridge",
    site_name == "venoco_bridge" ~ "Venoco Bridge",
    site_name == "other" ~ "Other"
  )) |> 
  # filter out Other observations
  filter(site_full != "Other")

```

```{r}
#| label: cleaning-water-parameters

# creating a new object from water parameters
water_parameters_clean <- water_parameters |> 
  # cleaning column names
  clean_names() |> 
  # "!" not operator, excluding unneeded columns
  select(!c(creator, editor)) |> 
  # renaming depth code to be elevation code to better represent methodology
  rename(elevation_code = depth_code) |> 
  ## Everything below came from visualizing the data first and defining what to fix/add
  # fixing a mistake in the data
  mutate(elevation_code = case_when(
    # anytime theres a 10 in elevation code, replace it with a 1
    elevation_code == 10 ~ 1,
    # everything else, keep the same
    .default = elevation_code)) |>
  mutate(
    # turning column into a factor
    elevation_code = as_factor(elevation_code),
    # defining the correct order for the levels of the factor
    elevation_code = fct_relevel(elevation_code,
                                 # write in the order you want
                                 "0", "1", "2", "3", "4", "5", "6"))
```


# 2. Joining 

## water quality = metadata + water parameters

```{r}
#| label: creating-water-quality-df



# full joining
water_quality <- full_join(
  # left data frame
  x = metadata_clean,
  # right data frame
  y = water_parameters_clean,
  # name the columns in each df that you want to join by (naming the shared column)
  by = c("global_id" = "parent_global_id")) |> 
  select(
    # information about the observation
    global_id, monitoring_datetime, time_of_wq_measurement, monitor_name, weather, site_name, other_site_name, site_full, x, y,
    # water information
    water_surface_elevation, flow, notes,
    #measurement information
    elevation_code, do_mg_l, do_percent_sat, conductivity_u_s_cm, salinity_ppt, temperature_c
  ) |> 
  # taking the date out of this column, dont want the time
  mutate(date = date(monitoring_datetime)) |> 
  # moving the new date column to a better location within the df
  relocate(date, .after = monitoring_datetime) |> 
  # dont need this column anymore
  select(!monitoring_datetime)
```

## water quality weather = water quality + weather

```{r}
#| label: join-water-quality-with-weather

# left join
water_quality_weather <- left_join(
  # left data frame
  x = water_quality,
  # right data frame
  y = weather_clean,
  # joining these dfs by their date column
  by = "date"
)
```

# 3. Wrangling 
## Dealing with Outliers in Salinity
```{r}
#| label: outlier

# creating new obj 
salinity_outlier_ids <- water_quality_weather |> 
  # filtering to only include outlier values
  filter(salinity_ppt > 1410) |> 
  # extracting the global id column as a vector usign pull
  pull(global_id)

```

## Get precipitation per month (site independent variable)
```{r}
#| label: prcp-monthly

monthly_prcp <- water_quality_weather |> 
  filter(elevation_code %in% c("0", "1", "2")) |> 
  filter(!site_name == "other") |> 
  distinct(month, year, wy, .keep_all = TRUE) |> 
  group_by(month, wy) |> 
  summarise(sum_prcp = sum(prcp, na.rm = TRUE)) |> 
  ungroup() |> 
  mutate(date = lubridate::make_date(.data$wy, .data$month, 1))


```



## Filtering Salinity & Summarizing Variables
```{r}
#| label: salinity-filtered

salinity_filtered <- water_quality_weather |> 
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # filtering the column global_id to exclude the values that match those in `salinity_outlier_ids`
  filter(!global_id %in% salinity_outlier_ids) |> 
  select(!other_site_name) |> 
  filter(!site_name == "other") |> 
  # setting the order of the levels (from lowest tp highest avg salinity value)
  # helps the order of the visualizations
  mutate(site_full = fct_relevel(site_full, 
                                 "Phelps Bridge", "East Channel", "Venoco Bridge")) |> 
  # taking out sites w/ NA
  filter(!site_full == "NA") |>
  # keeping only columns of interest (to reduce confusion)
  select(site_full, water_surface_elevation, elevation_code, salinity_ppt, temperature_c, month, year, wy) |> 
  # group by site full and month
  group_by(site_full, month, wy) |> 
  # calculate mean salinity (per month), and sum of precipitation (per month)
  summarize(mean_salinity = mean(salinity_ppt, na.rm = TRUE),
            .groups = "drop")



```

## Joining Monthly Precipitation (sum) w/ Monthly Salinity (avg.)

```{r}
#| label: joining-prcp

salinity_prcp <- salinity_filtered |> 
  left_join(monthly_prcp, by = c("month", "wy")) |> 
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
                        ordered = TRUE))


```



# 4. Comparing groups Viz

## Boxplot w/ Jitter
```{r}
#| label: boxplot

# Plot
salinity_prcp %>%
  ggplot( aes(x= site_full, y=mean_salinity, fill = site_full)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Distribution of Mean Salinity Values Across Sites") +
    xlab("Site") + 
  ylab("Mean Salinity (ppt)")


```


## Strip Chart w/ Standard Deviation
```{r}
#| label: strip-chart-std-dev

ggplot(data = salinity_prcp,
       mapping = aes(x = site_full,
                     y = mean_salinity)) +
  geom_jitter() +
  stat_summary(fun = mean, geom = "point", size = 3, color = "turquoise") +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1), geom = "errorbar", width = 0.2, color = "turquoise") +
  labs(x = "Site",
       y = "Mean Salinity (ppt)")


```

## Ridgeline Plot
```{r}
#| label: ridgeline

ggplot(salinity_prcp, aes(x = mean_salinity, y = site_full, fill = site_full)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none") +
  scale_fill_manual(values = natparks.pals("Banff")) +
  labs(x = "Salinity (ppt)",
       y = "Site",
       title= "Salinity Observation Distribution Across NCOS Sites")
 

```



# 5. Figure Visualizing


## Figure 1

```{r}
#| label: viz1

ggplot(data = salinity_prcp,
       mapping = aes(x = date,
                     y = sum_prcp)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Date",
       y = "Monthly Precipitation",
       title = "Monthly Precipitation (2020-2025)")

```


## Figure 2. - Not finished
```{r}
#| label: viz2

ggplot(data = salinity_filtered,
       mapping = aes(x = date,
                     y = water_surface_elevation)) +
  geom_point() +
  geom_line()

```

## Figure 3. - Not finished
```{r}
#| label: viz3

ggplot(data = salinity_filtered,
       mapping = aes(x = mean_salinity,
                     y = elevation_code)) +
  geom_col()


```

## Figure 4. - Not finished
```{r}
#| label: viz4

ggplot(data = salinity_filtered,
       mapping = aes(x = prcp,
                     y = salinity_ppt)) +
  geom_point() +
  facet_wrap(~ site_full)
  


```


