---
title: "Timeline Check-in"
author: "Madeline, Audrey, Maisie"
format: pdf
echo: false
toc: true
date: "2026-05-22"
execute: 
  message: false
  warning: false
---

```{r}
#| label: packages

# general use
library(tidyverse)
library(janitor)

# file organization
library(here)

# Boxplot Visualization

library(hrbrthemes)

library(viridis)

# text label instead of legend
library(cowplot)

# Ridgeline Plot
#install.packages("ggridges")
library(ggridges)
#install.packages("ggplot2")
library(ggplot2)

# Color Palette
library(NatParksPalettes)
library(wesanderson)

# Use ragg_png device for better font rendering and system font support
knitr::opts_chunk$set(dev = "ragg_png")

# to stack plots
library(patchwork)

```

```{r}
#| label: data

# weather data from NOAA weather station
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))

```

```{r}
#| label: all-figure-code

weather_clean <- weather |> 
  # made column names all lowercase
  clean_names() |> 
  # puts date in proper format and makes sure its understood as 
  # date (not as chr)
  mutate(date = mdy(date)) |> 
  # keeping only these 4 columns of interest
  select(date, prcp, tmax, tmin) |> 
  # creating new columns for month and year
  # extracting month from the date (needs to be in the standard date format 
  # for this function to work)
  mutate(month = month(date),
         year = year(date)) |> 
  mutate(wy = case_when(
    # when month os october or later, assign a value of year +1
    # in the water ywar for next year
    month >= 10 ~ year + 1,
    # keeping all months < 10 the same (sept or earlier), assigning value
    # of existing year
    .default = year
    ))


metadata_clean <- metadata |> 
  # col names easier to work with (all lowercase w/ _)
  clean_names() |> 
  # "!" is the "not-operator", columns we DONT want to keep, excluding
  select(!c(object_id, creator, editor)) |> 
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  rename(monitoring_datetime = monitoring_date) |> 
  mutate(month = month(monitoring_datetime),
         year = year(monitoring_datetime)) |> 
  # assigning water year
  mutate(wy = case_when(
    month >= 10 ~ year +1,
    .default = year )) |> 
  # creatig new column for site full name
  mutate(site_full = case_when(
    site_name == "east_channel" ~ "East Channel",
    site_name == "phelps_bridge" ~ "Phelps Bridge",
    site_name == "venoco_bridge" ~ "Venoco Bridge",
    site_name == "other" ~ "Other"
  )) |> 
  # filter out Other observations
  filter(site_full != "Other")

# creating a new object from water parameters
water_parameters_clean <- water_parameters |> 
  # cleaning column names
  clean_names() |> 
  # "!" not operator, excluding unneeded columns
  select(!c(creator, editor)) |> 
  # renaming depth code to be elevation code to better represent methodology
  rename(elevation_code = depth_code) |> 
  ## Everything below came from visualizing the data first and defining what 
  ## to fix/add
  # fixing a mistake in the data
  mutate(elevation_code = case_when(
    # anytime theres a 10 in elevation code, replace it with a 1
    elevation_code == 10 ~ 1,
    # everything else, keep the same
    .default = elevation_code)) |>
  mutate(
    # turning column into a factor
    elevation_code = as_factor(elevation_code),
    # defining the correct order for the levels of the factor
    elevation_code = fct_relevel(elevation_code,
                                 # write in the order you want
                                 "0", "1", "2", "3", "4", "5", "6"))

# full joining
water_quality <- full_join(
  # left data frame
  x = metadata_clean,
  # right data frame
  y = water_parameters_clean,
  # name the columns in each df that you want to join by 
  # (naming the shared column)
  by = c("global_id" = "parent_global_id")) |> 
  select(
    # information about the observation
    global_id, monitoring_datetime, time_of_wq_measurement, monitor_name,
    weather, 
    site_name, other_site_name, site_full, x, y,
    # water information
    water_surface_elevation, flow, notes,
    #measurement information
    elevation_code, do_mg_l, do_percent_sat, conductivity_u_s_cm,
    salinity_ppt, temperature_c
  ) |> 
  # taking the date out of this column, dont want the time
  mutate(date = date(monitoring_datetime)) |> 
  # moving the new date column to a better location within the df
  relocate(date, .after = monitoring_datetime) |> 
  # dont need this column anymore
  select(!monitoring_datetime)

# left join
water_quality_weather <- left_join(
  # left data frame
  x = water_quality,
  # right data frame
  y = weather_clean,
  # joining these dfs by their date column
  by = "date")

# creating new obj 
salinity_outlier_ids <- water_quality_weather |> 
  # filtering to only include outlier values
  filter(salinity_ppt > 1410) |> 
  # extracting the global id column as a vector usign pull
  pull(global_id)

# create new object from `water_quality_weather`
monthly_prcp <- water_quality_weather |> 
  # only include elevation codes 0, 1, and 2
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # get rid of site names w/ "other"
  filter(!site_full == "NA") |> 
  # removing any duplicate rows based on the combination of month, 
  # year, and wy
  distinct(month, year, prcp, site_full) |> 
  # grouping by month and year
  group_by(month, year) |> 
  # calculating the sum of prcp per month (not per site as it is a weather 
  # event and not site specific)
  summarise(sum_prcp = sum(prcp, na.rm = TRUE)) |> 
  # ungrouping
  ungroup() |> 
  # creating a date column to be read as a date
  mutate(date = lubridate::make_date(.data$year, .data$month, 1)) |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                   "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
                        ordered = TRUE))

salinity_filtered <- water_quality_weather |> 
  # keeping only elevation codes 0, 1, and 2
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # filtering the column global_id to exclude the values that match those
  # in `salinity_outlier_ids`
  filter(!global_id %in% salinity_outlier_ids) |> 
  # dropping the column for other_site_names
  select(!other_site_name) |> 
  # get rid of site names w/ "other"
  filter(!site_name == "other") |> 
  # setting the order of the levels (from lowest to highest avg salinity
  # value)
  # helps the order of the visualizations
  mutate(site_full = fct_relevel(site_full, 
                                 "Phelps Bridge", 
                                 "East Channel", 
                                 "Venoco Bridge")) |> 
  # taking out sites w/ NA
  filter(!site_full == "NA") |>
  # keeping only columns of interest (to reduce confusion)
  select(site_full, water_surface_elevation, elevation_code, salinity_ppt, 
         temperature_c, month, year, wy) |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May",
                                   "Jun", "Jul", "Aug", "Sep", "Oct",
                                   "Nov", "Dec"),
                        ordered = TRUE)) |> 
  # group by site full, month, and year
  group_by(site_full, month, year) |> 
  # calculate mean salinity (per site per month))
  summarize(mean_salinity = mean(salinity_ppt, na.rm = TRUE),
          # calculating mean temperature (in f) 
           mean_temp_f = (mean(temperature_c, na.rm = TRUE) * 9/5) + 32,
          .groups = "drop")

# creating new object ~salinity_prcp` 
salinity_prcp <- salinity_filtered |> 
  # joining `monthly_prcp` to `salinity_filtered` by the columns 
  # month and year
  left_join(monthly_prcp, by = c("month", "year"))

# create new object `exploratory_salinity` from `water_quality_weather`
exploratory_salinity <- water_quality_weather |> 
  # only keep elevation codes 0, 1, 2, and 3 (keeping 
  # three just for visualization)
  filter(elevation_code %in% c("0", "1", "2", "3")) |> 
  # filtering the column global_id to exclude the values that match
  #  those in `salinity_outlier_ids`
  filter(!global_id %in% salinity_outlier_ids) |> 
  # dropping the column `other_site_names`
  select(!other_site_name) |> 
  # filterign out site names that match "other"
  filter(!site_name == "other") |> 
  # setting the order of the levels (from lowest to highest avg
  # salinity value)
  # helps the order of the visualizations
  mutate(site_full = fct_relevel(site_full, 
                                 "Phelps Bridge", 
                                 "East Channel", 
                                 "Venoco Bridge")) |> 
  # taking out sites w/ NA
  filter(!site_full == "NA") |>
  # keeping only columns of interest (to reduce confusion)
  select(site_full, water_surface_elevation, elevation_code, salinity_ppt, 
         temperature_c, month, year) |>
  # group by site, elevation code, month, and year
  group_by(site_full, elevation_code, month, year) |> 
  # calculate mean salinity
  summarize(mean_salinity = mean(salinity_ppt, na.rm = TRUE),
            .groups = "drop") |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May",
                                   "Jun", "Jul", "Aug", "Sep", "Oct",
                                   "Nov", "Dec"),
                        ordered = TRUE))

# create a new object `monthly_surface_elv` from `water_quality_weather`
monthly_surfcae_elv <- water_quality_weather |> 
  # only include elevation codes 0, 1, and 2
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # Filter out site names matchign to "other"
  filter(!site_name == "other") |>  
  # group by site, month, and year
  group_by(site_full, month, year) |> 
  # calculate mean water surface elevation
  summarise(avg_elv = mean(water_surface_elevation, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a date column to be read as a date
  mutate(date = lubridate::make_date(.data$year, .data$month, 1)) |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May",
                                   "Jun", "Jul", "Aug", "Sep", "Oct",
                                   "Nov", "Dec"),
                        ordered = TRUE))

## Joining data frames 
elv_viz <- monthly_surfcae_elv |> 
  # joining by columns month, year, and site
  left_join(salinity_prcp,  by = c("month", "year", "site_full")) |> 
  # filtering out all NA average elevation values
  filter(!avg_elv == "NaN")

# creating a new object `precipitation_peaks` from `monthly_prcp`
precipitation_peaks <- monthly_prcp |>
  # including the dates of the top 4 monthly precipitation
  filter(date %in% c("2023-03-01", 
                     "2021-01-01", 
                     "2023-02-01", 
                     "2021-10-01")) |> 
  # ungrouping data frame
  ungroup() |> 
  # keeping these 4 distinct dates
  distinct(date, .keep_all = TRUE) |>
  # adding a label to each event
  mutate(label = c("Rain Event (6.05 in)",
                   "Rain Event (5.40 in)",  
                   "Rain Event (6.18 in)",  
                   "Rain Event (3.45 in)"),
         # setting the y position for each event
         label_y = c(6.5, 6.5, 8, 6.25)) |> 
  # sort chronologically (better legend appearance)
  arrange(date) |>                             
  # lock levels in that order
  mutate(label = factor(label, levels = label)) 


# creating a new object `precipitation_peaks_facet` from `precipitation_peaks`
precipitation_peaks_facet <- precipitation_peaks |>
  # duplicating all 4 peaks across each unique site in salinity_prcp
  cross_join(tibble(site_full = unique(salinity_prcp$site_full))) |> 
  # assigning y position for each peak label by date
  mutate(label_y_2 = case_when(
    # y position for January 2021 rain event
    date == "2021-01-01" ~ 40,
    # y position for February 2023 rain event
    date == "2023-02-01" ~ 35,
    # y position for March 2023 rain event
    date == "2023-03-01" ~ 45,
    # y position for October 2021 rain event
    date == "2021-10-01" ~ 50))



```





# Introduction


# Methods


# Results

```{r}
#| label: Figure 1
#| fig-height: 10
#| fig-width: 10

# Top half of Figure 1
# base layer: ggplot
# data: salinity_prcp
prcp_plot <- ggplot(data = monthly_prcp,
       # x axis: date
       mapping = aes(x = date,
                     # y axis: monthly precipitation sum
                     y = sum_prcp)) +
  # prcp displayed by columns
  geom_col(color = "darkblue") +
  # adding themeing 
  theme_bw() +
  # x-axis breaks every 6 months (to reduce cognitive load)
  # Format date axis labels as abbreviated month and full year
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  # angling labels to 45 degrees and to be right-aligned
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # expanding limits to make room for text
  scale_y_continuous(limits = c(0, 8)) + 
  # labeling x and y axes
  labs(x = "Date",
       y = "Total Monthly Precipitation (in)",
       # adding title
       title = "Total Monthly Precipitation at NCOS") + 
  # removing legend, redundant
  theme(legend.position = "none") +
  # adding in lines for prcp events
  geom_vline(data = precipitation_peaks,
             # centering them on dates
             aes(xintercept = date,
                 # coloring them by their established label
                 color = label),
             # signified by dotted lines
             linetype = "dotted",
             # width of lines
             linewidth = 1.2) +
  # adding custom color palette
  scale_color_manual(values = wes_palette("AsteroidCity1")) +
  # removing space between data and x-axis
  scale_y_continuous(expand = c(0,0.1)) +
  # adding labels to each dotted line from the `precipitation_peaks` object
  geom_text(data = precipitation_peaks,
            # x-axis location: date
            # y-axis location: set y value from earlier chunk
            # label = the defined label from earlier chunk
            # colro by label
            aes(x = date, y = label_y, label = label, color = label),
            # 45 degree angle for readability
            angle = 45,
            # slightly to right of vertical dotted line
            hjust = 1.1,
            # text sixe
            size = 3,
            # text bolding
            fontface = "bold")

# Bottom half of Figure1
# base layer ggplot
# data: salinity_prcp
salinity_plot <- ggplot(data = salinity_prcp,
       # x-axis: date
       mapping = aes(x = date,
                     # y-axis: mean_salinity
                     y = mean_salinity)) +
  # column displaying mean salinity
  geom_col(color = "grey") +
  # separating points from sites into different plots
  facet_wrap(~ site_full) +
  # adding themeing
  theme_bw() +
  # x breaks by year to reduce cognitive load, labeling by 
  # abbreviated month and full year
  scale_x_date(date_breaks = "12 months", date_labels = "%b %Y") +
  # angling x axis labels to 45 degrees and to be right-aligned
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # labeling x and y axes
  labs(x = "Date",
       y = "Monthly Average Salinity (ppt)",
       # adding title
       title = "Monthly Average NCOS Salinity w/ Precipitation Peaks",
       # adding legend title
       color = "Precipitation Events (in)") +
  # adding lines from the `precipitation_peaks_facet` object
  geom_vline(data = precipitation_peaks_facet,
             # oriented on the date
             aes(xintercept = date,
                 # different colors by label
                 color = label),
             # long dash for better viz
             linetype = "longdash",
             # line width
             linewidth = 0.75) +
  # custom colors
  scale_color_manual(values = wes_palette("AsteroidCity1")) +
  # removing space between data and x-axis
  scale_y_continuous(expand = c(0,0.1),
                     # defining vertical limit
                     limits = c(0, 65)) +
  # taking out legend (redundant when stacked)
  theme(legend.position = "none")

# Stacking plots
prcp_plot / salinity_plot +
  plot_annotation(title = "Figure 1. Precipitation Peaks at NCOS Overlaid on Monthly Average Salinity",
                  theme = theme(plot.title = element_text(face = "bold")))
```

_Figure 1. On the top, the total monthly precipitation is graphed by columns with the top 4 precipitation peaks depicted by colored dotted lines and labels. On the bottom, these precipitation peaks are overlaid onto the monthly average salinity for each of the three sites of intrest (Phelps Bridge, East Channel, and Venoco Bridge)._


```{r}
#| label: Figure 2
#| fig-height: 6
#| fig-width: 8

# base layer ggplot
# data: elevation visualization
ggplot(data = elv_viz,
       # x: average elevation
       mapping = aes(x = avg_elv,
                     # y: mean salinity
                     y = mean_salinity)) +
  # data shown by points
  geom_point(color = "maroon") +
  # each site gets own plot
  facet_wrap(~ site_full) +
  # custom color palette
  scale_color_manual(values = wes_palette("GrandBudapest1")) +
  # adding in themeing
  theme_bw() +
  # removing space between data and x-axis
  scale_y_continuous(expand = c(0,0.3)) +
  # labeling x and y axes 
  labs(x = "Monthly Average Water Surface Elevation (m)",
       y = "Monthly Average Salinity (ppt)",
       # adding title
       title = "Figure 2. How Salinity Relates to Water Surface 
Elevation (Venoco Bridge)") 


```

_Figure 2. This figure displays the relationship between monthly average water surface elevation and monthly average salinity as maroon points._


```{r}
#| label: Figure-3
#| fig-height: 6
#| fig-width: 8

# base layer: ggplot
# data: salinity_prcp
ggplot(data = salinity_prcp,
       # x axis: monthly average temperature in fahrenheit
       # y axis: monthly average salinity
       # color by site
       mapping = aes(x = mean_temp_f,
                     y = mean_salinity,
                     color = site_full)) +
  # data represented by points
  geom_point() +
  # each site gets its own plot
  facet_wrap(~ site_full) +
  # custom color palette
  scale_color_manual(values = wes_palette("Royal2")) +
  # removing space between data and x axis
  scale_y_continuous(expand = c(0, 0.2)) +
  # adding theming
  theme_bw() +
  # removing legend (redundant with facet labels)
  theme(legend.position = "none") +
  # labeling axes and title
  labs(x = "Monthly Average Temperature (°F)",
       y = "Monthly Average Salinity (ppt)",
       title = "Figure 3. How Temperature Relates to Salinity Across NCOS Sites")


```

_Figure 3. This figure displays the relationship between the monthly average temperature in degrees Fahrenheit and monthly average salinity for each of the 3 sites of interest (Phelps Bridge, East Channel, Venoco Bridge)._



# Discussion


# References 








