---
title: "Final Paper"
author: "Madeline, Audrey, Maisie"
format: pdf
echo: false
toc: true
date: "2026-06-10"
execute: 
  message: false
  warning: false
bibliography: references.bib
---

```{r}
#| label: packages

# general use
library(tidyverse)
library(janitor)

# file organization
library(here)

# Boxplot Visualization

library(hrbrthemes)

library(viridis)

# text label instead of legend
library(cowplot)

# Ridgeline Plot
#install.packages("ggridges")
library(ggridges)
#install.packages("ggplot2")
library(ggplot2)

# Color Palette
library(NatParksPalettes)
library(wesanderson)

# Use ragg_png device for better font rendering and system font support
knitr::opts_chunk$set(dev = "ragg_png")

# to stack plots
library(patchwork)

```

```{r}
#| label: data

# weather data from NOAA weather station
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))

```

```{r}
#| label: all-figure-code

weather_clean <- weather |> 
  # made column names all lowercase
  clean_names() |> 
  # puts date in proper format and makes sure its understood as 
  # date (not as chr)
  mutate(date = mdy(date)) |> 
  # keeping only these 4 columns of interest
  select(date, prcp, tmax, tmin) |> 
  # creating new columns for month and year
  # extracting month from the date (needs to be in the standard date format 
  # for this function to work)
  mutate(month = month(date),
         year = year(date)) |> 
  mutate(wy = case_when(
    # when month os october or later, assign a value of year +1
    # in the water ywar for next year
    month >= 10 ~ year + 1,
    # keeping all months < 10 the same (sept or earlier), assigning value
    # of existing year
    .default = year
    ))


metadata_clean <- metadata |> 
  # col names easier to work with (all lowercase w/ _)
  clean_names() |> 
  # "!" is the "not-operator", columns we DONT want to keep, excluding
  select(!c(object_id, creator, editor)) |> 
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  rename(monitoring_datetime = monitoring_date) |> 
  mutate(month = month(monitoring_datetime),
         year = year(monitoring_datetime)) |> 
  # assigning water year
  mutate(wy = case_when(
    month >= 10 ~ year +1,
    .default = year )) |> 
  # creatig new column for site full name
  mutate(site_full = case_when(
    site_name == "east_channel" ~ "East Channel",
    site_name == "phelps_bridge" ~ "Phelps Bridge",
    site_name == "venoco_bridge" ~ "Venoco Bridge",
    site_name == "other" ~ "Other"
  )) |> 
  # filter out Other observations
  filter(site_full != "Other")

# creating a new object from water parameters
water_parameters_clean <- water_parameters |> 
  # cleaning column names
  clean_names() |> 
  # "!" not operator, excluding unneeded columns
  select(!c(creator, editor)) |> 
  # renaming depth code to be elevation code to better represent methodology
  rename(elevation_code = depth_code) |> 
  ## Everything below came from visualizing the data first and defining what 
  ## to fix/add
  # fixing a mistake in the data
  mutate(elevation_code = case_when(
    # anytime theres a 10 in elevation code, replace it with a 1
    elevation_code == 10 ~ 1,
    # everything else, keep the same
    .default = elevation_code)) |>
  mutate(
    # turning column into a factor
    elevation_code = as_factor(elevation_code),
    # defining the correct order for the levels of the factor
    elevation_code = fct_relevel(elevation_code,
                                 # write in the order you want
                                 "0", "1", "2", "3", "4", "5", "6"))

# full joining
water_quality <- full_join(
  # left data frame
  x = metadata_clean,
  # right data frame
  y = water_parameters_clean,
  # name the columns in each df that you want to join by 
  # (naming the shared column)
  by = c("global_id" = "parent_global_id")) |> 
  select(
    # information about the observation
    global_id, monitoring_datetime, time_of_wq_measurement, monitor_name,
    weather, 
    site_name, other_site_name, site_full, x, y,
    # water information
    water_surface_elevation, flow, notes,
    #measurement information
    elevation_code, do_mg_l, do_percent_sat, conductivity_u_s_cm,
    salinity_ppt, temperature_c
  ) |> 
  # taking the date out of this column, dont want the time
  mutate(date = date(monitoring_datetime)) |> 
  # moving the new date column to a better location within the df
  relocate(date, .after = monitoring_datetime) |> 
  # dont need this column anymore
  select(!monitoring_datetime)

# left join
water_quality_weather <- left_join(
  # left data frame
  x = water_quality,
  # right data frame
  y = weather_clean,
  # joining these dfs by their date column
  by = "date")

# creating new obj 
salinity_outlier_ids <- water_quality_weather |> 
  # filtering to only include outlier values
  filter(salinity_ppt > 1410) |> 
  # extracting the global id column as a vector usign pull
  pull(global_id)

# create new object from `water_quality_weather`
monthly_prcp <- water_quality_weather |> 
  # only include elevation codes 0, 1, and 2
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # get rid of site names w/ "other"
  filter(!site_full == "NA") |> 
  # removing any duplicate rows based on the combination of month, 
  # year, and wy
  distinct(month, year, prcp, site_full) |> 
  # grouping by month and year
  group_by(month, year) |> 
  # calculating the sum of prcp per month (not per site as it is a weather 
  # event and not site specific)
  summarise(sum_prcp = sum(prcp, na.rm = TRUE)) |> 
  # ungrouping
  ungroup() |> 
  # creating a date column to be read as a date
  mutate(date = lubridate::make_date(.data$year, .data$month, 1)) |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                   "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
                        ordered = TRUE))

salinity_filtered <- water_quality_weather |> 
  # keeping only elevation codes 0, 1, and 2
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # filtering the column global_id to exclude the values that match those
  # in `salinity_outlier_ids`
  filter(!global_id %in% salinity_outlier_ids) |> 
  # dropping the column for other_site_names
  select(!other_site_name) |> 
  # get rid of site names w/ "other"
  filter(!site_name == "other") |> 
  # setting the order of the levels (from lowest to highest avg salinity
  # value)
  # helps the order of the visualizations
  mutate(site_full = fct_relevel(site_full, 
                                 "Phelps Bridge", 
                                 "East Channel", 
                                 "Venoco Bridge")) |> 
  # taking out sites w/ NA
  filter(!site_full == "NA") |>
  # keeping only columns of interest (to reduce confusion)
  select(site_full, water_surface_elevation, elevation_code, salinity_ppt, 
         temperature_c, month, year, wy) |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May",
                                   "Jun", "Jul", "Aug", "Sep", "Oct",
                                   "Nov", "Dec"),
                        ordered = TRUE)) |> 
  # group by site full, month, and year
  group_by(site_full, month, year) |> 
  # calculate mean salinity (per site per month))
  summarize(mean_salinity = mean(salinity_ppt, na.rm = TRUE),
          # calculating mean temperature (in f) 
           mean_temp_f = (mean(temperature_c, na.rm = TRUE) * 9/5) + 32,
          .groups = "drop")

# creating new object ~salinity_prcp` 
salinity_prcp <- salinity_filtered |> 
  # joining `monthly_prcp` to `salinity_filtered` by the columns 
  # month and year
  left_join(monthly_prcp, by = c("month", "year"))

# create new object `exploratory_salinity` from `water_quality_weather`
exploratory_salinity <- water_quality_weather |> 
  # only keep elevation codes 0, 1, 2, and 3 (keeping 
  # three just for visualization)
  filter(elevation_code %in% c("0", "1", "2", "3")) |> 
  # filtering the column global_id to exclude the values that match
  #  those in `salinity_outlier_ids`
  filter(!global_id %in% salinity_outlier_ids) |> 
  # dropping the column `other_site_names`
  select(!other_site_name) |> 
  # filterign out site names that match "other"
  filter(!site_name == "other") |> 
  # setting the order of the levels (from lowest to highest avg
  # salinity value)
  # helps the order of the visualizations
  mutate(site_full = fct_relevel(site_full, 
                                 "Phelps Bridge", 
                                 "East Channel", 
                                 "Venoco Bridge")) |> 
  # taking out sites w/ NA
  filter(!site_full == "NA") |>
  # keeping only columns of interest (to reduce confusion)
  select(site_full, water_surface_elevation, elevation_code, salinity_ppt, 
         temperature_c, month, year) |>
  # group by site, elevation code, month, and year
  group_by(site_full, elevation_code, month, year) |> 
  # calculate mean salinity
  summarize(mean_salinity = mean(salinity_ppt, na.rm = TRUE),
            .groups = "drop") |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May",
                                   "Jun", "Jul", "Aug", "Sep", "Oct",
                                   "Nov", "Dec"),
                        ordered = TRUE))

# create a new object `monthly_surface_elv` from `water_quality_weather`
monthly_surfcae_elv <- water_quality_weather |> 
  # only include elevation codes 0, 1, and 2
  filter(elevation_code %in% c("0", "1", "2")) |> 
  # Filter out site names matchign to "other"
  filter(!site_name == "other") |>  
  # group by site, month, and year
  group_by(site_full, month, year) |> 
  # calculate mean water surface elevation
  summarise(avg_elv = mean(water_surface_elevation, na.rm = TRUE)) |> 
  # ungroup data frame
  ungroup() |> 
  # create a date column to be read as a date
  mutate(date = lubridate::make_date(.data$year, .data$month, 1)) |> 
  # creating levels to `month`, putting them in a logical order
  mutate(month = factor(month,
                        levels = 1:12,
                        labels = c("Jan", "Feb", "Mar", "Apr", "May",
                                   "Jun", "Jul", "Aug", "Sep", "Oct",
                                   "Nov", "Dec"),
                        ordered = TRUE))

## Joining data frames 
elv_viz <- monthly_surfcae_elv |> 
  # joining by columns month, year, and site
  left_join(salinity_prcp,  by = c("month", "year", "site_full")) |> 
  # filtering out all NA average elevation values
  filter(!avg_elv == "NaN")

# creating a new object `precipitation_peaks` from `monthly_prcp`
precipitation_peaks <- monthly_prcp |>
  # including the dates of the top 4 monthly precipitation
  filter(date %in% c("2023-03-01", 
                     "2021-01-01", 
                     "2023-02-01", 
                     "2021-10-01")) |> 
  # ungrouping data frame
  ungroup() |> 
  # keeping these 4 distinct dates
  distinct(date, .keep_all = TRUE) |>
  # adding a label to each event
  mutate(label = c("Rain Event (6.05 in)",
                   "Rain Event (5.40 in)",  
                   "Rain Event (6.18 in)",  
                   "Rain Event (3.45 in)"),
         # setting the y position for each event
         label_y = c(6.5, 6.5, 8, 6.25)) |> 
  # sort chronologically (better legend appearance)
  arrange(date) |>                             
  # lock levels in that order
  mutate(label = factor(label, levels = label)) 


# creating a new object `precipitation_peaks_facet` from `precipitation_peaks`
precipitation_peaks_facet <- precipitation_peaks |>
  # duplicating all 4 peaks across each unique site in salinity_prcp
  cross_join(tibble(site_full = unique(salinity_prcp$site_full))) |> 
  # assigning y position for each peak label by date
  mutate(label_y_2 = case_when(
    # y position for January 2021 rain event
    date == "2021-01-01" ~ 40,
    # y position for February 2023 rain event
    date == "2023-02-01" ~ 35,
    # y position for March 2023 rain event
    date == "2023-03-01" ~ 45,
    # y position for October 2021 rain event
    date == "2021-10-01" ~ 50))



```

# Introduction

The University of California, Santa Barbara (UCSB) North Campus Open Space (NCOS) is a 136-acre restoration site located northwest of the main UCSB campus in Goleta, California. Prior to restoration, the site was occupied by Ocean Meadows Golf Course, the construction of which deposited fill soil that eliminated approximately half of the site's original water-holding capacity. This hydrological disruption degraded the natural wetland habitat and increased flood risk to surrounding homes. Following restoration efforts led by UCSB's Cheadle Center for Biodiversity and Ecological Restoration, the site received major recognition in September 2021 when FEMA issued a Letter of Map Revision (LOMR), formally removing the Flood Hazard Zone designation from nearby properties [@rickard2024].

The restored site now supports a diverse array of native species across its varying wetland habitats, including perennial grassland, salt marsh, freshwater creek, and restored vernal pools. Among the key factors differentiating these habitats is salinity, which varies considerably across the site. @clark2016 documented this variation in an early post-restoration water quality study, measuring salinity gradients alongside dissolved oxygen and nutrient levels to characterize overall wetland ecosystem health. Salinity levels at Venoco Bridge in particular were found to vary substantially in response to tidal influxes, rainfall, and freshwater inputs, with the greatest variation occurring around breach events connecting the slough to both freshwater and ocean sources.

This study investigates how changes in precipitation influence salinity levels across various sites at NCOS. It also explores how variables such as water elevation and temperature impact salinity. To explore this relationship, this study used the following data frames: “NOAA weather data”, “NCOS_YSI_Water_Quality_Monitoring”, and “YSI_Data_Begin,” to create visualizations to understand the broader environmental implications of this relationship. One application used from these data visualizations was observing changes in vegetation and how they relate to salinity, specifically through the elective map. Salinity is important to explore as it is known to have a strong impact on the biodiversity and species concentration within an ecosystem [@watt2006]. For example, salinity is known to have an impact on waterfowl breeding, as waterfowl typically like to breed in freshwater habitats, but exist in more brackish water after their breeding season. NCOS’s unique wetland ecosystem allows for both freshwater habitats and brackish water habitats, making it a perfect environment for healthy abundance of waterfowl. With the restored status of NCOS, tracking salinity and waterfowl together can depict the relationship between waterfowl abundance and salinity. Salinity is an important proxy in measuring the restoration and health of species at NCOS, and thus is extremely important for seeing the success of the project [@schacter2021]. This is because salinity has also been shown to potentially have a greater impact on ecosystems and its biota than other flows [@huckelbridge2010].

Through the following research questions, this study also explores the extent in which each of the factors affect salinity: How does salinity change with precipitation? How does salinity change with temperature? How does salinity change with water surface elevation?

# Methods

The data used in this analysis were collected through differing methods and by differing parties. Understanding the temporal and spatial coverage of these datasets is important for providing context about the exact location of our information, as well as for the proper summarization of variables in subsequent direct comparison.

Measurements of the variables salinity (ppt) and water temperature (℃), as well as conductivity and dissolved oxygen (not used for this analysis) were taken weekly by Cheadle Center student researchers at one foot depth intervals using a portable YSI Pro2030 at the three bridges that cross the upper slough: the Marsh trail bridges over the Phelps Creek outlet and “Dilling’s Link” across the east channel, and the Venoco access road bridge. From the bridges, the sensor is lowered to the water and data is recorded at the surface and at each foot of depth down to the bottom, or at each “elevation code”. Taking observations at each elevation code allows for the aspect of stratification to be taken into consideration. Or, in other words, the variation of these water parameters at different depths are important intricacies to understand when attempting to most accurately display the environmental conditions of this complex waterway. These variables are represented in the water parameters dataset ("YSI_Data_Begin_1.csv") and observations in this dataset span from September 3, 2020 to November 7, 2025 [@rickard2024].

The water surface elevation observations that acted as a secondary exploratory relationship to salinity were recorded at seven locations throughout water bodies of NCOS. One such location lies within our region of interest and a pressure transducer logger (YSI EXO1) is deployed on the north side of the water under Venoco Bridge. At a deployment elevation of 2.96 ft. NAVD88 (North American Vertical 50 Datum 1988), the levelogger takes water level measurements every 15 minutes. This ever-changing measurement allows researchers to contextualize changes within water composition and or concentration of solutes/particles?, providing for an increase in understanding of this complex system. This variable is represented in the metadata dataset ("NCOS_YSI_Water_Quality_Monitoring_0.csv") and observations in this dataset span from September 3, 2020 to November 7, 2025 [@rickard2024].

Precipitation observations, our main explanatory variable of interest within this analysis, were taken daily from the NOAA Santa Barbara 11 W, CA US station (Station ID: USW00053152) located just west of the Devereaux Lagoon (coordinates: 34°24'50.8"N 119°52'46.6"W). Rainfall information provides our analysis with a necessary external, environmental variable in our attempt to understand the drivers of salinity change. This variable is represented in the weather dataset ("NOAA-weather-data.csv") and observations in this dataset span from September 1, 2020 to November 30, 2025 [@NCEI_SB11W].

To characterize the structure and coverage of the dataset prior to inferential analysis, several descriptive visualizations were produced. Mean salinity values were summarized across the three monitoring sites, Venoco Bridge, East Channel, and Phelps Bridge, using boxplots to illustrate site-level variability of distribution. Observation counts were summarized by site and by elevation code using bar graphs to assess the variable coverage of available data across the sampling framework. Given the potential for non-normality in the salinity, water temperature, water surface elevation, and precipitation data, compounded by generally noted equipment malfunctions throughout the monitoring period [@rickard2024], Spearman's rank correlation tests were selected as the appropriate non-parametric measure of association. Spearman's correlation was applied to evaluate the monotonic relationships between salinity and each of these three environmental variables: water temperature, water surface elevation, and precipitation. A significance threshold of α = 0.05 was applied to all tests.

# Results

## Data Characteization

Observation counts varied across the three monitoring sites, as visualized by Figure 1, with Venoco Bridge contributing the greatest number of records (n \> 600), followed by Phelps Bridge (n \> 375) and East Channel (n \> 300). This imbalance in representation should be considered when interpreting site-level comparisons. Observations were further unevenly distributed across elevation codes, as seen in Figure 2, with codes 0 and 1 accounting for the majority of records and code 2 being notably underrepresented across all sites. This limited representation at deeper elevation codes constrains the reliability of any patterns observed at those levels.

```{r}
#| label: observation-per-site
#| fig-height: 6
#| fig-width: 8

# base layer: ggplot
# data: water_quality_weather, filtered and releveled inline
ggplot(data = water_quality_weather |> 
         # removing outlier IDs, "other" sites, and keeping only
         #elevation codes 0, 1, and 2
         filter(!global_id %in% salinity_outlier_ids,
                site_name != "other",
                elevation_code %in% c("0", "1", "2")) |> 
         # setting site order from lowest to highest average salinity
         mutate(site_full = fct_relevel(site_full,
                                        "Phelps Bridge", 
                                        "East Channel", 
                                        "Venoco Bridge")),
       # x axis: site name
       mapping = aes(x = site_full)) +
  # custom fill colors
  scale_fill_manual(values = natparks.pals("Volcanoes")) +
  # removing the space between data and x axis
  scale_y_continuous(expand = c(0, 0.1),
                     # definign limits to include all data
                     limits = c(0, 700)) +
  # bar chart with fill colored by site
  geom_bar(aes(fill = site_full)) +
  # adding theming
  theme_bw() +
  # removing legend (redundant)
  theme(legend.position = "none") +
  # labeling x and y axes
  labs(x = "Site",
       y = "Number of Observations",
       # adding title
       title = "Figure 1. Number of Observations at Each Site")

```

*Figure 1. The number of observations taken at each of the three sites (Phelps Bridge, East Channel, and Venoco Bridge) is displayed as a bar graph. Data from @rickard2024.*

```{r}
#| label: observation-per-code
#| fig-height: 6
#| fig-width: 8

# base layer: ggplot
# data: water_quality_weather, filtered inline
ggplot(data = water_quality_weather |> 
         # removing outlier IDs, "other" sites, and keeping only
         #elevation codes 0, 1, and 2
         filter(!global_id %in% salinity_outlier_ids,
                site_name != "other",
                elevation_code %in% c("0", "1", "2")),
       # x axis: elevation code
       mapping = aes(x = elevation_code)) +
  # bar chart with fill colored by elevation code
  geom_bar(aes(fill = elevation_code)) +
  # custom fill colors
  scale_fill_manual(values = natparks.pals("Banff")) +
  # removing space between data and x axis
  scale_y_continuous(expand = c(0, 0.1),
                     # defining vertical limits 
                     limits = c(0, 650)) +
  # adding theming
  theme_bw() +
  # removing legend (redundant)
  theme(legend.position = "none") +
  # labeling x and y axes
  labs(x = "Elevation Code",
       y = "Number of Observations",
       # adding title
       title = "Figure 2. Observation Count Across Elevation Codes")

```

*Figure 2. The number of observations taken at each elevation code (0, 1, and 2) during the study period is displayed as a bar graph. Data from @rickard2024.*

Mean salinity differed substantially across the three sites (Figure 3). Venoco Bridge exhibited the highest mean salinity at approximately 25 ppt, followed by East Channel at approximately 19 ppt, and Phelps Bridge at approximately 2.5 ppt. The interquartile ranges at Venoco Bridge and East Channel were comparably sized, indicating similar degrees of variability at those two sites, while Phelps Bridge displayed a notably compressed distribution, suggesting more consistently low and stable salinity conditions. This makes contextual sense as the Phelp’s Bridge site is noted to remain at freshwater levels quite consistently because of a lack of direct tidal influence [@rickard2024].

```{r}
#| label: salinity-boxplot
#| fig-height: 6
#| fig-width: 8

  # base layer: ggplot
  # data: salinity~prcp
  ggplot(data = salinity_prcp,
         # x: site name, y: mean salinity, coloring in by site
         aes(x= site_full, y=mean_salinity, fill = site_full)) +
  # boxplot to display data  
  geom_boxplot(color = "black") +
  # adding jitter to points for better visualization
  geom_jitter(color="black", size=0.4, alpha=0.9, height = 0) +
  # custom coloring and a 0.6 transparency to visualize data points
  scale_fill_manual(values = alpha(wes_palette("GrandBudapest1"), 0.6)) +
  # removing space between data and x axis
  scale_y_continuous(expand = c(0, 0.1)) +
  # adding in a theme
  theme_bw() +
  # taking out legend
  theme(legend.position="none") +
  # labeling x and y axes
  labs(x = "Site",
       y = "Mean Salinity (ppt)",
       # adding in a title
       title = "Figure 3. Distribution of Mean Salinity Values Across Sites")


```

*Figure 3. Mean salinity values across the study period are displayed usign boxplots for each of the three sites (Phelps Bridge, East Channel, and Venoco Bridge). Data from @rickard2024.*

When salinity was examined by elevation code (Figure 4), East Channel and Venoco Bridge appeared to have lower mean salinity at elevation code 2 than at codes 0 and 1. While higher salinity at lower elevation codes might be expected given that denser saline water tends to underlie fresher water in coastal systems through saltwater intrusion dynamics [@mazhar2022], the observed inverse pattern at East Channel and Venoco Bridge is likely an artifact of the limited number of observations at elevation code 2 rather than a true reflection of salinity conditions at that elevation.

```{r}
#| label: salinity-by-code
#| fig-height: 6
#| fig-width: 8

# base layer: ggplot
# data: `exploratory_salinity`
ggplot(data = exploratory_salinity,
       # x: elevation code
       mapping = aes(x = elevation_code,
                     # y: mean salinity
                     y = mean_salinity)) +
  # data represented by open points
  geom_point(shape = 21) +
  # calculating median of data & representing by points
  stat_summary(fun = median,
               geom = "point",
               # custom color
               color = "maroon") +
  # calculating median of data, connecting medians per site by a line
  stat_summary(fun = median,
               geom = "line",
               # custom color
               color = "maroon",
               # all data into one group to connect across x categories
               group = 1) +
  # labeling x and y axes
  labs(x = "Elevation Code",
       y = "Mean Salinity (ppt)",
       # adding in title
       title = "Figure 4. Mean Salinity Differences Between 
Elevation Codes") +
  # each site has own plot
  facet_wrap(~ site_full) +
  # adding custom themeing
  theme_bw() +
  # defining the discrete elevation codes
  scale_x_discrete(breaks = c(0,1, 2, 3)) +
  # Moves the y-axis tick marks and labels to the right
  # removing space between y axis and data
  scale_y_continuous(position = "right",
                     expand = c(0,0.3)) +
  # Flips the coordinate system
  coord_flip()


```

*Figure 4. The mean salinity values for different elevation codes (0, 1, 2, and 3) are displayed for each of the three sites (Phelps Bridge, East Channel, and Venoco Bridge) across the study period. Data from @rickard2024.*

## Precipitation and Salinity Relationship

```{r}
#| label: prcp
#| fig-height: 10
#| fig-width: 10

# Top half of Figure 1
# base layer: ggplot
# data: salinity_prcp
prcp_plot <- ggplot(data = monthly_prcp,
       # x axis: date
       mapping = aes(x = date,
                     # y axis: monthly precipitation sum
                     y = sum_prcp)) +
  # prcp displayed by columns
  geom_col(color = "darkblue") +
  # adding themeing 
  theme_bw() +
  # x-axis breaks every 6 months (to reduce cognitive load)
  # Format date axis labels as abbreviated month and full year
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  # angling labels to 45 degrees and to be right-aligned
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  # expanding limits to make room for text
  scale_y_continuous(limits = c(0, 8)) + 
  # labeling x and y axes
  labs(x = "Date",
       y = "Total Monthly Precipitation (in)",
       # adding title
       title = "Total Monthly Precipitation at NCOS") + 
  # removing legend, redundant
  theme(legend.position = "none") +
  # adding in lines for prcp events
  geom_vline(data = precipitation_peaks,
             # centering them on dates
             aes(xintercept = date,
                 # coloring them by their established label
                 color = label),
             # signified by dotted lines
             linetype = "dotted",
             # width of lines
             linewidth = 1.2) +
  # adding custom color palette
  scale_color_manual(values = wes_palette("AsteroidCity1")) +
  # removing space between data and x-axis
  scale_y_continuous(expand = c(0,0.1)) +
  # adding labels to each dotted line from the `precipitation_peaks` object
  geom_text(data = precipitation_peaks,
            # x-axis location: date
            # y-axis location: set y value from earlier chunk
            # label = the defined label from earlier chunk
            # colro by label
            aes(x = date, y = label_y, label = label, color = label),
            # 45 degree angle for readability
            angle = 45,
            # slightly to right of vertical dotted line
            hjust = 1.1,
            # text sixe
            size = 3,
            # text bolding
            fontface = "bold")

# Bottom half of Figure1
# base layer ggplot
# data: salinity_prcp
salinity_plot <- ggplot(data = salinity_prcp,
       # x-axis: date
       mapping = aes(x = date,
                     # y-axis: mean_salinity
                     y = mean_salinity)) +
  # column displaying mean salinity
  geom_col(color = "grey") +
  # separating points from sites into different plots
  facet_wrap(~ site_full) +
  # adding themeing
  theme_bw() +
  # x breaks by year to reduce cognitive load, labeling by 
  # abbreviated month and full year
  scale_x_date(date_breaks = "12 months", date_labels = "%b %Y") +
  # angling x axis labels to 45 degrees and to be right-aligned
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  # labeling x and y axes
  labs(x = "Date",
       y = "Monthly Average Salinity (ppt)",
       # adding title
       title = "Monthly Average NCOS Salinity w/ Precipitation Peaks",
       # adding legend title
       color = "Precipitation Events (in)") +
  # adding lines from the `precipitation_peaks_facet` object
  geom_vline(data = precipitation_peaks_facet,
             # oriented on the date
             aes(xintercept = date,
                 # different colors by label
                 color = label),
             # long dash for better viz
             linetype = "longdash",
             # line width
             linewidth = 0.75) +
  # custom colors
  scale_color_manual(values = wes_palette("AsteroidCity1")) +
  # removing space between data and x-axis
  scale_y_continuous(expand = c(0,0.1),
                     # defining vertical limit
                     limits = c(0, 65)) +
  # taking out legend (redundant when stacked)
  theme(legend.position = "none")

# Stacking plots
prcp_plot / salinity_plot +
  plot_annotation(title = "Figure 5. Precipitation Peaks at NCOS Overlaid on Monthly Average Salinity",
                  theme = theme(plot.title = element_text(face = "bold")))
```

*Figure 5. On the top, the total monthly precipitation is graphed by columns with the top 4 precipitation peaks depicted by colored dotted lines and labels. On the bottom, these precipitation peaks are overlaid onto the monthly average salinity for each of the three sites of intrest (Phelps Bridge, East Channel, and Venoco Bridge). Data from @rickard2024.*

Monthly total precipitation across the study period was dominated by four notable peaks occurring in January 2021, October 2021, February 2023, and March 2023, with comparatively low rainfall recorded throughout the remainder of the record (Figure 5). When these precipitation peaks were overlaid onto monthly average salinity at each site (Figure 5), no consistent or widespread salinity response was apparent across all three sites. Isolated instances of elevated salinity coinciding with precipitation peaks were observed at East Channel and Venoco Bridge, though these patterns were not consistent enough to suggest a direct or reliable relationship as peak salinity periods were not always accompanied by a precipitation peak event. Phelps Bridge showed little discernible response to precipitation events likely because of the low recorded salinity at this particular site, making patterns difficult to observe. The relationship's scatterplot (Figure 6) displays a seemingly lacking correlation between these two variables with a heavy majority of points congregating in the low precipitation, low salinity corner (Phelps Bridge) and by the vertical pattern of points displaying that low precipitation levels can and do give rise to a large range in monthly average salinity values.

```{r}
#| label: salinity-prcp
#| fig-height: 6
#| fig-width: 8

# base layer: ggplot
# data: `salinity_prcp`
ggplot(data = salinity_prcp,
       # x: sum precipitation
       mapping = aes(x = sum_prcp,
                     # y: mean salinity
                     y = mean_salinity,
                     # each site own color
                     color = site_full)) +
  # data represented by points
  geom_point() +
  # each site has own plot
  facet_wrap(~ site_full) +
  # custom color palette
  scale_color_manual(values = wes_palette("Royal2")) +
  # removing space between data and x axis
  scale_y_continuous(expand = c(0,0.1)) +
  # removing space between data and y axis
  scale_x_continuous(expand = c(0, 0.1)) +
  # custom themeing
  theme_bw() +
  # removing legend (redundant)
  theme(legend.position = "none") +
  # labeling x and y axes
  labs(x = "Monthly Precipitation (in)",
       y = "Monthly Average Salinity (ppt)",
       # adding figure title
       title = "Figure 6. How Precipitation Relates to Salinity Across NCOS Sites")

```

*Figure 6. This figure displays the relationship between the monthly total precipitation in inches and the monthly average salinity in ppt. The figure is divided and colored by observations from each of the three sites of interest (Phelps Bridge, East Channel, and Venoco Bridge). Data from @rickard2024.*

The results of the Spearman's rank correlation test support this same observation, indicating no statistically significant relationship between monthly total precipitation and monthly average salinity (rho = 0.024, S = 932,653, p = 0.746) (Figure 4). The near-zero rho suggests negligible association between these variables at the monthly timescale, which may reflect lag effects between precipitation events and salinity response, or the dominant influence of other drivers such as tidal exchange operating independently of local precipitation inputs. The high p-value (p = 0.746) indicates that the observed correlation was well within the range of what would be expected by chance alone if no true relationship existed, providing no statistical evidence of an association between monthly precipitation and monthly mean salinity at this scale.

## Water Surface Elevation and Salinity Relationship

Figure 7 displays the relationship between monthly average water surface elevation and monthly average salinity. While no strong or obvious linear relationship is apparent, the cluster of points exhibits a subtle positive tendency, where lower monthly average water surface elevation corresponds generally to lower monthly average salinity, and higher water surface elevations are associated with somewhat higher salinity values.

```{r}
#| label: water-surface-elevation-salinity
#| fig-height: 6
#| fig-width: 8

# base layer ggplot
# data: elevation visualization
ggplot(data = elv_viz,
       # x: average elevation
       mapping = aes(x = avg_elv,
                     # y: mean salinity
                     y = mean_salinity)) +
  # data shown by points
  geom_point(color = "maroon") +
  # each site gets own plot
  facet_wrap(~ site_full) +
  # custom color palette
  scale_color_manual(values = wes_palette("GrandBudapest1")) +
  # adding in themeing
  theme_bw() +
  # removing space between data and x-axis
  scale_y_continuous(expand = c(0,0.3)) +
  # labeling x and y axes 
  labs(x = "Monthly Average Water Surface Elevation (m)",
       y = "Monthly Average Salinity (ppt)",
       # adding title
       title = "Figure 7. How Salinity Relates to Water Surface 
Elevation (Venoco Bridge)") 


```

*Figure 7. This figure displays the relationship between monthly average water surface elevation and monthly average salinity as maroon points. Data from @rickard2024.*

No statistically significant relationship was identified between monthly average water surface elevation and monthly average salinity (rho = 0.228, S = 30,660, p = 0.075) (Figure 2). The p-value of 0.075 indicates that, assuming no true relationship existed, there would be a 7.5% probability of observing a correlation as strong as or stronger than the one obtained. While this narrowly exceeds the significance threshold of α = 0.05, it suggests the result is not easily dismissed and may reflect a weak but potentially meaningful positive correlation between water surface elevation and salinity that warrants further investigation with a larger sample. This pattern may reflect tidal dynamics or episodic saltwater intrusion events, whereby higher water levels correspond to greater marine influence within the system, rather than the effect of increased precipitation events.

## Temperature and Salinity Relationship

Figure 8 displays the relationship between monthly average water temperature and monthly average salinity across the three monitoring sites. At Phelps Bridge, the majority of points are tightly clustered at low salinity values (below 5 ppt) across the full range of observed temperatures, with only a small number of higher salinity observations scattered across the temperature range, suggesting no directional relationship between the two variables at this site. At East Channel and Venoco Bridge, salinity values are more widely distributed across the y-axis but similarly show no consistent directional pattern with increasing temperature, high and low salinity values appear across the full range of observed temperatures at both sites.

```{r}
#| label: temp-salinity
#| fig-height: 6
#| fig-width: 8

# base layer: ggplot
# data: salinity_prcp
ggplot(data = salinity_prcp,
       # x axis: monthly average temperature in fahrenheit
       # y axis: monthly average salinity
       # color by site
       mapping = aes(x = mean_temp_f,
                     y = mean_salinity,
                     color = site_full)) +
  # data represented by points
  geom_point() +
  # each site gets its own plot
  facet_wrap(~ site_full) +
  # custom color palette
  scale_color_manual(values = wes_palette("Royal2")) +
  # removing space between data and x axis
  scale_y_continuous(expand = c(0, 0.2)) +
  # adding theming
  theme_bw() +
  # removing legend (redundant with facet labels)
  theme(legend.position = "none") +
  # labeling axes and title
  labs(x = "Monthly Average Water Temperature (°F)",
       y = "Monthly Average Salinity (ppt)",
       title = "Figure 8. How Temperature Relates to Salinity Across NCOS Sites")


```

*Figure 8. This figure displays the relationship between the monthly average temperature in degrees Fahrenheit and monthly average salinity for each of the three sites of interest (Phelps Bridge, East Channel, Venoco Bridge). Data from @rickard2024.*

The Spearman's correlation test confirmed no statistically significant relationship between monthly average temperature and salinity (rho = 0.032, p = 0.666). The p-value of 0.666 indicates that the observed correlation was well within the range of what would be expected by chance alone if no true relationship existed, providing no statistical evidence of a meaningful association between these variables at the monthly temporal scale. The near-zero rho further reinforces the negligible nature of this association, though indirect effects of temperature on salinity mediated through processes such as evaporation may not be adequately captured at this resolution.

# Discussion

Salinity levels at NCOS exhibited notable spatial variation across the three monitoring sites, as illustrated in Figure 3. Venoco Bridge consistently displayed the highest mean salinity values at approximately 25 ppt, followed by East Channel at approximately 19 ppt, and Phelps Bridge at approximately 2.5 ppt. This site-level variation is consistent with the findings of @watson2009, who demonstrated that salinity concentrations vary across coastal wetland systems and that this spatial variability has meaningful consequences for species distribution and plant community composition. @galliari2021 further support this pattern, finding that salinity in coastal wetlands is shaped by several interacting natural processes including tidal flooding, evaporation, rainfall, and salts dissolving from sediment, and that salinity can differ substantially within a single wetland depending on elevation, groundwater flow, and tidal inundation frequency. The spatial variation observed across NCOS sites is therefore likely a reflection of differences in tidal connectivity, elevation, and proximity to marine influence across the three locations rather than the influence of any single measurable variable.

## Research Question 1: How salinity relates to precipitation

As shown in Figures 1 and 6, monthly total precipitation and monthly average salinity displayed no consistent or directional relationship across the study period. The Spearman's rank correlation test confirmed this observation, returning a near-zero rho value (0.024) and a p-value of 0.746, indicating that the observed correlation was well within the range of what would be expected by chance alone if no true relationship existed. These findings are consistent with @nordio2022, who found that while rainfall caused short-term increases in groundwater level and short-term decreases in salinity through freshwater dilution, precipitation did not exert a strong influence on salinity over longer periods such as weeks or months, the timescale of our study. Instead, salinity was more closely tied to groundwater levels, evapotranspiration, and the movement of groundwater between terrestrial and aquatic zones [@nordio2022]. This is particularly relevant to our study given that salinity was observed at a weekly timescale and further summarized at a monthly timescale, which likely obscures any short-term dilution responses to individual precipitation events. @clark2016 further supports this interpretation in the context of NCOS specifically, noting that while rainfall lowered salinity by adding freshwater to Devereux Slough, major storm and breach events reconnected the slough to the ocean, allowing seawater to enter and raise salinity again, effectively counteracting any freshwater dilution signal at longer timescales. The relative proximity of Venoco Bridge to the slough and thus the Pacific Ocean supports the findings seen in Figure 3 that Venoco Bridge sees the highest salinity values, followed by East Channel and Phelps Bridge. The hydrology of Devereux Slough is also highly variable across years. @rickard2024 documents that the sand berm at the mouth of the slough typically breaches once per year, temporarily making the system tidal, though in particularly wet years such as 2023 the berm breached three times due to frequent and intense storms. These breach events represent episodic, large-magnitude inputs of saline ocean water that likely have a far greater influence on salinity than background precipitation levels, and are not captured by a monthly precipitation variable alone. Examining the relationship between mean salinity values and slough breach events, or the cumulative number of days the system remained tidal, would be a logical next step toward a deeper understanding of salinity drivers at NCOS. The sustained tidal connectivity associated with breach events likely explains salinity variability more effectively than episodic precipitation inputs, and capturing this temporal dynamic would provide a more ecologically meaningful characterization of the forces shaping salinity within the slough.

## Research Question 2: How salinity relates to water surface elevation

Figure 7 displays the relationship between monthly average water surface elevation and monthly average salinity. While no strong linear pattern is apparent, the distribution of points exhibits a subtle positive tendency, where lower water surface elevations correspond generally to lower salinity values and higher elevations are associated with somewhat higher salinity. The Spearman's correlation test returned a moderately positive rho (rho = 0.228, p = 0.075), which narrowly missed the significance threshold of α = 0.05. While this result does not constitute statistical significance under the criteria applied in this analysis, it should not be easily dismissed. The p-value of 0.075 indicates that assuming no true relationship existed, there would be only a 7.5% probability of observing a correlation as strong as this one by chance, suggesting a weak but potentially meaningful tendency that warrants further investigation with a larger sample and perhaps a finer temporal coverage. While originally thought to be a proxy for precipitation, water surface elevation at NCOS specifically seems to be more telling of, and perhaps representative of, the occurrences of tidal breaches between Devereaux Slough and the Pacific Ocean; higher water surface elevations correspond to greater marine influence within the system and thus elevated salinity. @clark2016 noted that salinity at NCOS often increases with depth and that breach events reconnecting the slough to the ocean were associated with marked increases in salinity, further supporting the idea that water surface elevation and tidal connectivity are meaningfully linked to salinity dynamics at this site.

## Research Question 3: How salinity relates to temperature

As shown in Figure 8, there is no consistent directional relationship between monthly average water temperature and monthly average salinity was apparent across any of the three monitoring sites. At Phelps Bridge, salinity values remained tightly clustered near zero across the full range of observed temperatures, while at East Channel and Venoco Bridge, salinity values were broadly distributed across the y-axis with no clear pattern of increase or decrease with temperature. The Spearman's correlation test confirmed no statistically significant relationship between these variables (rho = 0.032, p = 0.666), with the p-value indicating that the observed correlation was well within the range expected by chance alone under the null hypothesis of no association. While salinity is known to fluctuate seasonally in coastal wetland systems, the Mediterranean climate of Santa Barbara limits the degree to which temperature alone drives meaningful seasonal contrasts in salinity. @anderson2011 provides useful context here, finding that saltwater intrusion in tidal freshwater wetlands was most pronounced during periods of low river flow in late summer and fall, when sea levels were relatively higher and tidal connectivity was stronger. This suggests that while temperature may serve as a proxy for seasonality, the salinity changes associated with seasonal shifts at NCOS are more directly driven by tidal dynamics and the alternation between wet and dry periods than by temperature itself [@wood2014]. The indirect pathways through which temperature might influence salinity, such as elevated evaporation rates during warmer months concentrating salts in the water column, are unlikely to be captured at the summarized monthly temporal resolution used in this analysis, and may warrant examination at finer timescales in future work.

## Data Challenges and Limitations

Several challenges inherent to collecting and working with environmental monitoring data in coastal wetland systems are relevant to the interpretation of these results. The physical nature of data collection at NCOS presents practical limitations worth considering. As all three monitoring sites are located on bridges over open water, consistent sensor placement is difficult to guarantee across sampling events. Small differences in how far a sensor is lowered into the water column, or where laterally it sits relative to the current, could introduce variability in readings that is unrelated to true changes in salinity. The structure of the bridges themselves may also locally disturb water flow, meaning readings taken immediately adjacent to bridge pilings may not fully represent conditions across the broader channel or collection site. Additionally, sensors consistently utilized in estuarine environments are prone to buildup of algae and other organisms over time, which can reduce measurement accuracy between maintenance visits. Perhaps most critically, the conditions that are emerging as the most ecologically relevant to this study, large storm events and slough breaches, are also the conditions most likely to make bridge access unsafe or impractical, potentially resulting in data gaps at the very moments when salinity dynamics are most actively changing. This is further reflected in the monitoring record itself, as @rickard2024 notes that equipment malfunction was experienced to some extent across all years of the study period, resulting in data gaps that may have influenced the representativeness of monthly averages and reduced statistical power in the correlation analyses.

The monthly temporal resolution of this dataset represents a meaningful constraint on the kinds of relationships that can be detected. As @nordio2022 found, the effects of precipitation on salinity are most pronounced at short timescales (hours to days) through freshwater dilution of surface and groundwater. Summarizing data at the monthly scale smooths over these short-term dynamics and likely contributes to the weak and non-significant correlation observed between precipitation and salinity in this study. Daily resolution data would better capture the salinity responses to individual storm events and breach dynamics that are known to occur at NCOS.

Beyond the variables examined in this study, several additional factors are known to influence salinity in coastal wetland systems and were not directly captured in this dataset. Tidal exchange is among the most significant drivers of salinity in systems like Devereux Slough, where the breaching of the sand berm at the mouth of the slough introduces large pulses of ocean water that can rapidly elevate salinity across the system [@clark2016; @rickard2024]. The timing, frequency, and duration of these breach events varied considerably across the study period, from a single breach in dry years to three breach events in the exceptionally wet year of 2023, representing a source of salinity variability that a monthly precipitation variable alone cannot adequately represent. Evaporation is another important unmeasured driver, particularly during dry summer months when the concentration of salts in standing water can increase substantially in the absence of freshwater inputs [@clark2016]. Groundwater flow, which @galliari2021 identifies as a key control on salinity in coastal wetlands, also operates independently of surface precipitation and at timescales that are not easily captured by monthly monitoring data. Finally, the year-to-year hydrological variability documented by @rickard2024, ranging from drought years with minimal rainfall and single large storm events to exceptionally wet years with nearly twice the average annual precipitation, represents a form of temporal variation that complicates the detection of stable relationships between environmental variables and salinity across the full study period.

## Synthesis

Taken together, the results of this study reflect the inherent complexity of salinity dynamics in coastal wetland systems. No single environmental variable, precipitation, temperature, or water surface elevation, was found to be a statistically significant predictor of monthly average salinity at NCOS, which is consistent with the understanding that salinity in systems like Devereux Slough is shaped by the simultaneous interaction of multiple hydrological, physical, and biological processes operating across different timescales [@galliari2021]. The consistent site-level differences in mean salinity observed across Venoco Bridge, East Channel, and Phelps Bridge suggest that local factors such as proximity to tidal inputs and degree of hydrological connectivity play a more dominant role in structuring salinity patterns than any single measured variable at the monthly scale. As @watson2009 demonstrate, this spatial variability in salinity has meaningful ecological consequences, with different salinity regimes supporting different plant communities and shifts in salinity potentially negatively affecting biodiversity [@amores2013], vegetation cover [@huckelbridge2010], carbon sequestration, and microbial activity [@zhao2017]. Continued long-term monitoring at NCOS, ideally at finer temporal resolution and inclusive of tidal flux and groundwater measurements, would substantially improve the ability to isolate the relative contributions of these interacting drivers and better characterize the ecological implications of salinity variability across the site.

# References
